*DCVtestkit*: a R package for linearity assessment and analysis of quality control dilution curves

### 2022-11-01 Manuscript/manuscript.qmd
Fixed some citation issues.

Linearity assessment is one of the criteria used to evaluate the precision of a quantitative measurement procedure as seen in calibration studies @rodríguez1993  and drug analysis @needleman1990. In metabolomics/lipidomics analysis, dilution curves are plotted for each molecule measured from quality control sample, generated by pooling equal aliquots from each study sample. Molecules that exhibits a non-linear relationship are rejected before further statistical analysis, as their quantification in different sample groups might be compromised. Dilution curves are usually plotted using general-purpose software like Excel with their corresponding Pearson Correlation Coefficient value. A threshold value of 0.8 is used to balance the risk of accepting or removing non-linear signals related to relevant molecules.

@@

Next, summary statistics are calculated via `summarise_dilution_table` for each dilution curve. Besides the Pearson Correlation Coefficient, additional parameters introduced are results from Mandel’s Fitting Test @andrade2013 in which a low $p$ value gives sufficient evidence that a quadratic model fits better than a linear model, indicating the curve may not be linear. Another parameter is Percent Residual Accuracy @logue2018 which ranges from $-\infty$ to $100$. If the curve is linear, the value should be close to $100$. The software also calculates the concavity of a fitted quadratic model to identify if the curve is dominantly non-linear at high (concavity $<0$) or low (concavity $>0$) concentrations.

>>>>

Linearity assessment is one of the criteria used to evaluate the precision of a quantitative measurement procedure as seen in calibration studies from @rodríguez1993  and drug analysis from @needleman1990. In metabolomics/lipidomics analysis, dilution curves are plotted for each molecule measured from quality control sample, generated by pooling equal aliquots from each study sample. Molecules that exhibits a non-linear relationship are rejected before further statistical analysis, as their quantification in different sample groups might be compromised. Dilution curves are usually plotted using general-purpose software like Excel with their corresponding Pearson Correlation Coefficient value. A threshold value of 0.8 is used to balance the risk of accepting or removing non-linear signals related to relevant molecules.

@@

Next, summary statistics are calculated via `summarise_dilution_table` for each dilution curve. Besides the Pearson Correlation Coefficient, additional parameters introduced are Mandel’s Fitting Test results from @andrade2013 in which a low $p$ value gives sufficient evidence that a quadratic model fits better than a linear model, indicating the curve may not be linear. Another parameter is Percent Residual Accuracy from @logue2018 which ranges from $-\infty$ to $100$. If the curve is linear, the value should be close to $100$. The software also calculates the concavity of a fitted quadratic model to identify if the curve is dominantly non-linear at high (concavity $<0$) or low (concavity $>0$) concentrations.
